---
title: "SAL413_FinalProject"
author: "Jared Weber"
date: "2024-11-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Could a model that just observes that the over under and moneyline of a game predict fantasy defense scores better than ESPN's projections?

```{r}
library(tidyverse)
library(nflfastR)
library(nflreadr)
library(vtable)
```

loads in fantasy and betting datsets that filters bye weeks

```{r}
game_info <- load_schedules(seasons = 2019:2023)

fantasy_info <- read_csv("espnff_player_projections_v2.csv") %>% 
  filter(default_position == "DEF") %>% 
  filter(!(projected_points == 0 & actual_points == 0)) %>% 
  rename("week" = "game_week", "team" = "nfl_team_abbr", "espn_projected_points" = "projected_points")
```

win probability calculation

```{r}
game_bet_info <- game_info %>% 
  select(season, week, away_team, home_team, away_moneyline, home_moneyline, total_line) %>% 
  mutate(away_implied_win_prob_juiced = ifelse(away_moneyline <= -100,
                                               1 + 100/(away_moneyline - 100),
                                               100/(away_moneyline + 100)),
         home_implied_win_prob_juiced = ifelse(home_moneyline <= -100,
                                               1 + 100/(home_moneyline - 100),
                                               100/(home_moneyline + 100)),
         away_implied_win_prob = away_implied_win_prob_juiced/
           (away_implied_win_prob_juiced + home_implied_win_prob_juiced),
         home_implied_win_prob = home_implied_win_prob_juiced/
           (away_implied_win_prob_juiced + home_implied_win_prob_juiced))
```

organize game data into team data

```{r}
game_bet_info_away <- game_bet_info %>% 
  select(season, week, away_team, total_line, away_implied_win_prob) %>% 
  rename("team" = "away_team","implied_win_prob" = "away_implied_win_prob")

game_bet_info_home <- game_bet_info %>% 
  select(season, week, home_team, total_line, home_implied_win_prob)%>% 
  rename("team" = "home_team","implied_win_prob" = "home_implied_win_prob")

team_bet_info <- rbind(game_bet_info_away, game_bet_info_home) %>% 
  mutate(team = ifelse(team == "OAK", "LV", team)) %>% 
  mutate(team = ifelse(team == "LA", "LAR", team))
```

join data sets and filter into model dataset

```{r}
fantasy_bet_info <- inner_join(fantasy_info, team_bet_info, by = c('season', 'week', 'team'))

fantasy_bet_info_2019_2022 <- fantasy_bet_info %>% 
  filter(season %in% c(2019:2022))
```

preliminary scatter plots

```{r}
ggplot(fantasy_bet_info, aes(x = implied_win_prob, y = actual_points)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, size = 1) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma)

ggplot(fantasy_bet_info, aes(x = total_line, y = actual_points)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, size = 1) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma)
```

```{r}
fantasy_lm = lm(actual_points ~ implied_win_prob + 
                  total_line + factor(season) + factor(team), 
                data = fantasy_bet_info_2019_2022)
summary(fantasy_lm)
```

builds projection calucation from linear model into test dataset

```{r}
fantasy_bet_info_2023 <- fantasy_bet_info %>% 
  filter(season == 2023) %>% 
  mutate(our_projected_points = round(fantasy_lm$coefficients["(Intercept)"] +
           fantasy_lm$coefficients["implied_win_prob"] * implied_win_prob +
           fantasy_lm$coefficients["total_line"] * total_line, 1))
```

```{r}
# Function to calculate evaluation metrics
evaluate_projections <- function(actual, projected) {
  data.frame(
    MAE = mean(abs(actual - projected)),
    RMSE = sqrt(mean((actual - projected)^2))
  )
}
```

mae and rmse are better for us but not significantly different

```{r}
# Compare projections
comparison <- bind_rows(
  Projection_1 = evaluate_projections(fantasy_bet_info_2023$actual_points, 
                                      fantasy_bet_info_2023$espn_projected_points),
  Projection_2 = evaluate_projections(fantasy_bet_info_2023$actual_points, 
                                      fantasy_bet_info_2023$our_projected_points),
  .id = "Projection"
)
```

```{r}
st(fantasy_bet_info_2023)
```

our distribution is consistently a few points higher than ESPN

```{r}
# Reshape the data into long format
fantasy_bet_info_2023_long <- fantasy_bet_info_2023 %>%
  select(espn_projected_points, our_projected_points) %>% 
  pivot_longer(cols = everything(), names_to = "Defense Fantasy Points", values_to = "Projected Fantasy Points")

# Create the density plot
ggplot(fantasy_bet_info_2023_long, aes(x = `Projected Fantasy Points`, 
                                       color = `Defense Fantasy Points`, fill = `Defense Fantasy Points`)) +
  geom_density(alpha = 0.4) +  # alpha controls transparency
  labs(
    title = "Density Plot of Defense Fantasy Point Projection",
    x = "Projected Fantasy Points",
    y = "Density"
  ) +
  theme_minimal()
```


```{r}
# Reshape the data into long format
fantasy_bet_info_2023_long <- fantasy_bet_info_2023 %>%
  select(espn_projected_points, our_projected_points, actual_points) %>% 
  pivot_longer(cols = everything(), names_to = "Defense Fantasy Points", values_to = "Fantasy Points")

# Create the density plot
ggplot(fantasy_bet_info_2023_long, aes(x = `Fantasy Points`, 
                                       color = `Defense Fantasy Points`, fill = `Defense Fantasy Points`)) +
  geom_density(alpha = 0.4) +  # alpha controls transparency
  labs(
    title = "Density Plot of Defense Fantasy Point Projection",
    x = "Fantasy Points",
    y = "Density"
  ) +
  theme_minimal()
```

```{r}
ggplot(fantasy_bet_info_2023, aes(x = espn_projected_points, y = actual_points)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, size = 1) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma)

ggplot(fantasy_bet_info_2023, aes(x = our_projected_points, y = actual_points)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, size = 1) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma)
```


We over predict more often, they under predict more often

```{r}
# Residual density plot
residuals <- fantasy_bet_info_2023 %>%
  mutate(
    residual_1 = actual_points - espn_projected_points,
    residual_2 = actual_points - our_projected_points
  )

ggplot(residuals) +
  geom_density(aes(x = residual_1, fill = "ESPN Projection"), alpha = 0.5) +
  geom_density(aes(x = residual_2, fill = "Our Projection"), alpha = 0.5) +
  labs(title = "Density of Fantasy Point Residuals", x = "Fantasy Point Residuals", y = "Density") +
  scale_fill_manual(values = c("ESPN Projection" = "blue", "Our Projection" = "red", "Avg Projection" = "green")) +
  theme_minimal()
```

average residual for each calculation, ESPN has a more consistent underpredicting bias

```{r}
fantasy_bet_info_2023 %>%
  mutate(
    residual_1 = mean(espn_projected_points - actual_points),
    residual_2 = mean(our_projected_points - actual_points)
  ) %>%
  summarise(
    ESPN_Bias_Projection = mean(residual_1),
    Our_Bias_Projection = mean(residual_2)
  )
```

Our projections are not that significantly better at projecting fantasy scores but they do not have a consistent over or under prediction issue

```{r}
t.test(abs(fantasy_bet_info_2023$projected_points - fantasy_bet_info_2023$actual_points),
       abs(fantasy_bet_info_2023$our_projected_points - fantasy_bet_info_2023$actual_points),
       paired = TRUE)
```
